Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-18 ieee_paper.tex
changes.

For a total of $N$ points, there will be $N$ linear systems to solve,
each of size $n_z \times n_z$. Each linear solve produces a row of the
RBF-FD differentiation matrix $D_{n_z}$, resulting in a $N \times N$
matrix with $n_zN$ nonzero entries. To evaluate the derivative at all
points in the domain, one multiplies the source vector $f$ by the 
derivative matrix $D_{n_z}$ to obtain the vector $g$ of derivative values, 
$$
  g = D_{n_z} f
$$
computation of a single derivative has been reduced to a SpMV, 
where each row has $n_z$ nonzeros. In
practice, $n_z=32$ in two-dimensional flows and 64 or 100 for
three-dimensional flows. These numbers are on a par with what is used in
finite-element codes. Notice that the nonzero elements of the matrix
correspond to a relation of nearest neighbor. This relationship is not
necessarily symmetric as shown in Figure~\ref{fig:rbf_stencils} which
implies a non-symmetric sparsity pattern in the derivative matrix.

@@

Many problems in fluid dynamics and in the geosciences require the
solution to transport equations of the form
$$
\pf{Q}{t} = f(Q,Q_x, Q_y, Q_z, \Laplacian{Q})
$$
where $Q$ is a vector of unknowns %(e.g., 3D Navier-Stokes has
%three components of velocity, pressure and temperature). 
For example, in solving a system of equations, it is
often necessary to compute the $x$ derivative of multiple functions,
typically four for the Euler equations, or five for the Navier-Stokes
equations. Multiple right-hand sides transform a SpMV into a SpMM
(Sparse Matrix/dense Matrix multiplication), which improves register
utilization and decreases cache misses by vectorizing over the
multiple source vectors. Further improvements are possible by
recognizing that different derivative matrices (e.g., $x$, $y$, $z$ 
and Laplacian operators), 
have the same sparsity distribution as long as the derivative stencils
remain unchanged; only the values in the sparse matrix change. 
change.  Thus, rather than computing a derivative of multiple
functions, one can calculate multiple derivatives of a single
function. The increased memory bandwidth due to an increase in the
number of derivative matrices is offset by better cache utilization,
leading to an overall performance benefit. In practice however, the number of 
vectors whose derivatives are required is limited, limiting the 
performance of the matrix/vector multiplication. Similarly, the number
of different derivatives in a particular computation is limited. It is 
for this reason that we seek to combine the benefits of multiple
vectors and multiple matrices simultaneously. We limit this initial 
study to four vectors and four matrices.

\todo{Add few sentence with discretization}

@@

In term of memory layout, we adopt a format similar to the ELLPACK
format to store the matrix since the number of nonzeros per row is
constant throughout the matrix. The matrix is given in two arrays. The
first one is the classical {\tt col\_id} array that describes the
sparsity structure of the matrix. Each row of {\tt col\_id} consists of $n_z$
entries that label the column index of each nonzero in$A$. The
other array is the {\tt data} array that gives the values of the 
nonzeros in the four matrices. The values of the matrices are interleaved
so that the value $A^l_{col\_id[k]}$ of the $k$th nonzero of the
$l$th matrix is at index $(k-1)*4+l-1$. The vectors {\tt x} are 
interleaved so that {\tt x}$^l_i$ and {\tt x}$^{l+1}_i$ are
consecutive in memory. The vector {\tt y} is similarly
interleaved. This is schematically shown in Figure~\ref{fig:mat_struct}, 
where the elements of $y$ are grouped in sets of 16, which correspond
to four derivatives applied to four functions, at a single point.

@@

\begin{figure}[bth]
  \begin{center}
    \scalebox{.9}{
      \begin{tabular}{|c|l|}
        %\hline
        %& & \\
        \hline
        $b_i$ & number of bytes per index \\
        $b_x$ & number of bytes per value \\
        $n_z$ & number of nonzeros per row $A$ \\
        $n_r$ & number of column/rows of $A$ \\
        $n_c$ & total number of nonzeros\\
        $n_v$ & number of {\tt x} vectors \\
        $n_m$ & number of matrices \\
        $s_M$ & size of the $n_m$ matrices in bytes\\
        $s_x$ & size of the $n_v$ {\tt x} vectors in bytes\\
        $s_y$ & size of the $n_v n_m$ {\tt y} vectors in bytes\\
        \hline
        $cl$    & size of a cache line in bytes\\
        $b_{wT}$ & number of bytes written to memory  \\
        $b_{rT}$ & minimum number of bytes read from memory  \\
        $b_T$   & minimum number of bytes transferred \\
        $B_{rT}$ & maximum number of bytes read from memory \\
        $B_T$   & maximum number of bytes transferred \\
        $O$     & number of floating point operations \\
        $I_b$   & maximum computational intensity\\
        $I_w$   & minimum computational intensity\\        
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{Notation relative to the matrices (upper section) and to the benchmarks (lower section.)}
  \label{tab:not}
\end{figure}

@@

\begin{figure*}
  \centering 
  \subfigure[Absolute performance ]{\includegraphics[width=.49\linewidth]{figures/gflops_peak.pdf}\label{fig:gflops_peak_perf}}
%
  \subfigure[Relative performance to using one matrix and one
    vector.]{\includegraphics[width=.49\linewidth]{figures/speedup_wrt_base.pdf}\label{fig:speedup}}

@@

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of four lanes (sometimes called channels) of 128 bits, 
which are made of four segments of 32 bits. A swizzling operation allows
the reorder of the elements within each lane, while a permutation allows 
changes the order of entire lanes. 
Figure~\ref{fig:swizzling_permutation}
illustrates these two possibilities. Both permutation and swizzling
support common reordering and broadcast. In the code, the ``permute''
function needs to use the {\tt \_mm512\_permute4f128\_epi32}
instruction which is only defined on integers and therefore in place typecasts
are necessary to convert to a vector to/from the integer type.
\todo{I added a comment stating we preferred swizzling.}
(Note that we prefer swizzling to permutation if possible because we believe 
it is cheaper. Many instructions have swizzling capability, while permuation
is its own instruction. \todo{reference?}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/swizzling_lanes.pdf}
  \caption{Swizzling enables the reorder or broadcast of elements within
    each 128 bit group. Permutation enables to reorder or broadcast
    whole 128 bit groups. All lanes must be sizzeled identically.}
  \label{fig:swizzling_permutation}
\end{figure}
\todo{I added to the caption.}

Every operation that loads data from memory into a register
transfers 512 bits, or 16 floats at a time. A single nonzero
in the matrix will cause 16 operations to be performed, but it only
uses 4 floats from the matrix and 4 floats from the {\tt x} vector. So
we would like to execute a single instruction to load the data for four 
nonzeros in the matrix, representing 16 floats from the matrix and 16 floats from the vectors
at once. (The number of nonzeros on a row in our application is
always a multiple of 4; if it weren't, we would add some explicit
zeroes in the data structure for padding.)
\todo{ reread paragraph above. single instruction to load 16 floats. Confused.}

@@

Now that $A_h^1,A_h^2,A_h^3,A_h^4$ and $x_m^1,x_m^2,x_m^3,x_m^4$ are
properly laid out in the vectors, a partial value for $y_i$ can be
computed. Once all the nonzeros of the row have been processed, the
exact $y_i$ value can be sent to memory using a the No Read No Global
Order hint to optimize write traffic on the memory bus.
\todo{ added "write" to traffic }

\section{Experimental Validation}
\label{sec:expe}

@@

The different matrices used to perform our experiments and analysis
are presented in Figure~\ref{fig:spy_plots}. The
Supercompact matrices (Figure~\ref{fig:supercompact}) have been
generated to only have nonzero elements in the first 32 columns of
the matrix. That way we ensure the simplest form of multiplication,
the number of elements from the {\tt x} vector data transfered from
the memory is be minimum, they are transferred efficiently and remain 
in cache. The Compact matrices
(Figure~\ref{fig:compact}) are generated to have 32 nonzeros per row
centered around the diagonal and it represents the ideal case for many
applications that rely on sparse matrix vector multiplication. The nonzero
elements of the Random
matrices (Figure~\ref{fig:compact}) see their nonzero elements
randomly (uniformly) distributed in the matrix; they represent the
worst case scenario for a cache based architecture where the cache
reutilisation is the lowest from one row to the next.

The other two types of matrices are used in derivative stencils in 2D
and 3D RBF-FD calculations (Figure~\ref{fig:rbf2dnorcm} pictures a
$32$-points stencil of a $23^2$ 2D grid and
Figure~\ref{fig:rbf2dnorcm} shows a $32$-points stencil of a $8^3$
grid). We also apply a Cuthill-Mckee reordering to all the matrices. This
ordering technique is designed to reduce the distance between the 
nonzeros and the diagonal, hopefully allowing a better cache
reutilisation. The reordered version of these two matrices can be seen
in Figure~\ref{fig:rbf2drcm} and in Figure~\ref{fig:rbf3drcm}.

@@

\begin{figure*}
\centering
  \subfigure[
    Entries with "cpp" denote cases where
    coding was performed without MIC vector
    instructions.]{\includegraphics[width=.40\linewidth]{figures/bandwidth_read_write.pdf}\label{fig:band_rw}}
%
  \hspace{.3in}
%
  \subfigure[
    The speed on the CPU matches the speed with vector instructions.
    All memory is aligned on 32
    bytes.]{\includegraphics[width=.40\linewidth]{figures/host_test1_readwrite_no_temporal_hint.pdf}\label{fig:read_write}}

@@

\begin{figure*}
  \centering
  \subfigure[Performance of $y=Ax$ on the MIC. Squares: base 1/1 case,
    solid circles: 4/4 case implemented in C++, solid triangles: 4/4
    case implemented with MIC vector instructions. $-O3$ compilation
    options. Each symbol has two cases: with (red) and without blue) 
    with and without Reverse Cuthill-McKee. The grid is $96^3$.]
    {\includegraphics[width=.40\linewidth]{figures/mic_performance_nb_threads.pdf}\label{fig:perf_mic}}
%
\hspace{.4in}
%
  \subfigure[Performance on the host on a sparse matrix with $64^3$ rows. Manual
    vectorization brings significant improvement there as
    well.]{\includegraphics[width=.40\linewidth]{figures/64x64x64.pdf}\label{fig:perf_host}}
  \caption{Vectorization matters in practice. Differences in BW
    translate in different performance.\todo{this two plots should ``correlate'' }}
\end{figure*}

\begin{figure*}[t]
  \centering
  \subfigure[Supercompact and random.]{\includegraphics[width=.48\linewidth]{figures/random_supercompact.pdf}}
  \subfigure[Compact and RBF-FD]{\includegraphics[width=.48\linewidth]{figures/rbf_compact.pdf}}

  %\subfigure[Supercompact]{\includegraphics[width=.48\linewidth]{figures/supercompact_max_perf.pdf}}
  %\subfigure[Compact]{\includegraphics[width=.48\linewidth]{figures/compact_max_perf.pdf}} \\
  %\subfigure[Random]{\includegraphics[width=.48\linewidth]{figures/random_max_perf.pdf}}
  %\subfigure[RBF 3D]{\includegraphics[width=.48\linewidth]{figures/rbf_max_perf_run1.pdf}}
  \caption{Performance of "best code" accross instance types.}
\end{figure*}

>>>>

For a total of $N$ points, there will be $N$ linear systems to solve,
each of size $n_z \times n_z$. Each linear solve produces a row of the
RBF-FD differentiation matrix $A_{n_z}$, resulting in a $N \times N$
matrix with $n_zN$ nonzero entries. To evaluate the derivative at all
points in the domain, one multiplies the source vector $x$ by the 
derivative matrix $A_{n_z}$ to obtain the vector $y$ of derivative values, 
$$
  y = A_{n_z} x
$$
computation of a single derivative has been reduced to a SpMV, 
where each row has $n_z$ nonzeros\footnote{We use non-conventional 
symbology for differentiation matrix, source and destination vectors
anticipating what follows.}. 
In practice, $n_z=32$ in two-dimensional flows and 64 or 100 for
three-dimensional flows. These numbers are on a par with what is used in
finite-element codes. Notice that the nonzero elements of the matrix
correspond to a relation of nearest neighbor. This relationship is not
necessarily symmetric as shown in Figure~\ref{fig:rbf_stencils} which
implies a non-symmetric sparsity pattern in the derivative matrix.

@@

Many problems in fluid dynamics and in the geosciences require the
solution to transport equations of the form
\begin{equation}
\pf{Q}{t} = f(Q,\grad{Q},\Laplacian{Q})  \label{eq:Q}
\end{equation}
where $Q$ is a vector of unknowns %(e.g., 3D Navier-Stokes has
%three components of velocity, pressure and temperature). 
For example, in solving a system of equations, it is
often necessary to compute the derivative of multiple functions along
the same coordinate direction,
typically four for the Euler equations, or five for the Navier-Stokes
equations. Multiple right-hand sides transform a SpMV into a SpMM
(Sparse Matrix/dense Matrix multiplication), which improves register
utilization and decreases cache misses by vectorizing over the
multiple source vectors. Further improvements are possible by
recognizing that different derivative matrices (e.g., gradient and 
Laplacian operators), 
have the same sparsity distribution as long as the derivative stencils
remain unchanged; only the values in the sparse matrix change. 
change.  Thus, rather than computing a derivative of multiple
functions, one can calculate multiple derivatives of a single
function. The increased memory bandwidth due to an increase in the
number of derivative matrices is offset by better cache utilization,
leading to an overall performance benefit. In practice however, the number of 
vectors whose derivatives are required is limited, limiting the 
performance of the matrix/vector multiplication. Similarly, the number
of different derivatives in a particular computation is limited. It is 
for this reason that we seek to combine the benefits of multiple
vectors and multiple matrices simultaneously. We limit this initial 
study to four vectors and four matrices. 
\todo{I have remove reference to $x$, $y$, ... to avoid confusion with what follows.}

\todo{Add few sentence with discretization. Check Erik}
In discrete form, Equation (\ref{eq:Q}) takes the form
$$
  y^{i,j} = A^i x^j 
$$
where $j$ ranges over the number of source vectors, $i$ ranges of the number of derivative
matrices and there are $i*j$ output vectors $y$.

@@

In term of memory layout, we adopt a format similar to the ELLPACK
format to store the matrix since the number of nonzeros per row is
constant throughout the matrix. The matrix is given in two arrays. The
first one is the classical {\tt col\_id} array that describes the
sparsity structure of the matrix. Each row of {\tt col\_id} consists of $n_z$
entries that label the column index of each nonzero in $A$. The
other array is the {\tt data} array that gives the values of the 
nonzeros in the four matrices. The values of the matrices are interleaved
so that the value $A^l_{col\_id[k]}$ of the $k$th nonzero of the
$l$th matrix is at index $(k-1)*4+l-1$. The vectors {\tt x} are 
interleaved so that {\tt x}$^l_i$ and {\tt x}$^{l+1}_i$ are
consecutive in memory. The vector {\tt y} is similarly
interleaved. This is schematically shown in Figure~\ref{fig:mat_struct}, 
where the elements of $y$ are grouped in sets of 16, which correspond
to four derivatives applied to four functions, at a single point.

@@

\begin{figure}[bth]
  \begin{center}
    \scalebox{.9}{
      \begin{tabular}{|c|l|}
        %\hline
        %& & \\
        \hline
        $b_i$ & number of bytes per index \\
        $b_x$ & number of bytes per value \\
        $n_z$ & number of nonzeros per row of $A$ \\
        $n_r$ & number of column/rows of $A$ \\
        $n_c$ & total number of nonzeros\\
        $n_v$ & number of {\tt x} vectors \\
        $n_m$ & number of matrices \\
        $s_M$ & size of the $n_m$ matrices in bytes\\
        $s_x$ & size of the $n_v$ {\tt x} vectors in bytes\\
        $s_y$ & size of the $n_v n_m$ {\tt y} vectors in bytes\\
        \hline
        $cl$    & size of a cache line in bytes\\
        $b_{wT}$ & number of bytes written to memory  \\
        $b_{rT}$ & minimum number of bytes read from memory  \\
        $b_T$   & minimum number of bytes transferred \\
        $B_{rT}$ & maximum number of bytes read from memory \\
        $B_T$   & maximum number of bytes transferred \\
        $O$     & number of floating point operations \\
        $I_b$   & maximum computational intensity\\
        $I_w$   & minimum computational intensity\\        
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{Notation relative to the matrices (upper section) and to the benchmarks (lower section.)}
  \label{tab:not}
\end{figure}

@@

\begin{figure*}
  \centering 
  \subfigure[Absolute performance ]{\includegraphics[width=.49\linewidth]{figures/gflops_peak.pdf}\label{fig:gflops-peak-perf}}
%
  \subfigure[Relative performance to using one matrix and one
    vector.]{\includegraphics[width=.49\linewidth]{figures/speedup_wrt_base.pdf}\label{fig:speedup}}

@@

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of four lanes (sometimes called channels) of 128 bits, 
which are made of four segments of 32 bits. A swizzling operation allows
the reorder of the elements within each lane, while a permutation reorders
entire lanes, without changing their contents. 
Figure~\ref{fig:swizzling_permutation}
illustrates these two possibilities. Both permutation and swizzling
support common reordering and broadcast. In the code, the ``permute''
function needs to use the {\tt \_mm512\_permute4f128\_epi32}
instruction which is only defined on integers and therefore in place typecasts
are necessary to convert to a vector to/from the integer type.
\todo{I added a comment stating we preferred swizzling.}
(Note that we prefer swizzling to permutation if possible because we believe 
it is cheaper. Many instructions have swizzling capability, while permuation
is its own instruction. \todo{do we need a reference to my note?})

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/swizzling_lanes.pdf}
  \caption{Swizzling enables the reorder or broadcast of elements within
    each 128 bit group. Permutation enables to reorder or broadcast
    whole 128 bit groups. All lanes must be sizzeled identically.
    \todo{I added to the caption.}}
  \label{fig:swizzling_permutation}
\end{figure}

Every operation that loads data from memory into a register
transfers 512 bits, or 16 floats at a time. A single nonzero
in the matrix will cause 16 operations to be performed, but it only
uses 4 floats from the matrix and 4 floats from the {\tt x} vector. So
we would like to execute a single instruction to load the data for four 
nonzeros in the matrix, representing 16 floats from the matrix and 16 floats from the vectors
at once. (The number of nonzeros on a row in our application is
always a multiple of 4; if it weren't, we would add some explicit
zeroes in the data structure for padding.)
\todo{Erik, reread paragraph above. single instruction to load 16 floats. Confused.}

@@

Now that $A_h^1,A_h^2,A_h^3,A_h^4$ and $x_m^1,x_m^2,x_m^3,x_m^4$ are
properly laid out in the vectors, a partial value for $y_i$ can be
computed. Once all the nonzeros of the row have been processed, the
exact $y_i$ value can be sent to memory using a the No Read No Global
Order hint to optimize write traffic on the memory bus.
\todo{I added "write" to traffic }

\section{Experimental Validation}
\label{sec:expe}
In this section, we describe a series of numerical experiments to confirm 
the performance predictions of Section~\ref{sec:model}. To better
understand our results, we first perform some bandwidth measurements, followed
by the actual SPmV.  \todo{We might be criticized for not comparing with other work.}

@@

The different matrices used to perform our experiments and analysis
are presented in Figure~\ref{fig:spy_plots}. The
Supercompact matrices (Figure~\ref{fig:supercompact}) have been
generated to only have nonzero elements in the first 32 columns of
the matrix. As a result, only 32 values of {\tt x} are read
from memory, and the expense associated with cache misses and gather
operation is removed from consideration. The cost associated with 
{\tt A} remains.   \todo{reread the above.}
The Compact matrices
(Figure~\ref{fig:compact}) are generated to have 32 nonzeros per row
centered around the diagonal and it represents the ideal case for many
applications that rely on sparse matrix vector multiplication. The nonzero
elements of the Random
matrices (Figure~\ref{fig:compact}) see their nonzero elements
randomly (uniformly) distributed in the matrix; they represent the
worst case scenario for a cache based architecture where the cache
reutilisation is the lowest from one row to the next.

The other two types of matrices are used in derivative stencils in 2D
and 3D RBF-FD calculations (Figure~\ref{fig:rbf2dnorcm} pictures a
$32$-points stencil of a $23^2$ 2D grid and
Figure~\ref{fig:rbf2dnorcm} shows a $32$-points stencil of a $8^3$
grid). We also apply a Cuthill-Mckee reordering to all the matrices. 
The 2D and 3D grid sizes are chosen to have comparable matrix sizes. One notes
that the bandwidth reduction is higher in the 2D configuration, which leads
to better performance.  \todo{we do not show performance results in 2D. Do we
need to change anything? Do we have a reference for RCM?} 
This ordering technique is designed to reduce the distance between the 
nonzeros and the diagonal, hopefully allowing a better cache
reutilisation. The reordered version of these two matrices can be seen
in Figure~\ref{fig:rbf2drcm} and in Figure~\ref{fig:rbf3drcm}.

@@

\begin{figure*}
\centering
  \subfigure[Run on the MIC with 240 threads, 64-byte memory alignment.]{\includegraphics[width=.40\linewidth]{figures/bandwidth_read_write.pdf}\label{fig:band_rw}}
%
  \hspace{.3in}
%
  \subfigure[
    Run on the host with 32 threads, 32-byte memory alignment.]{\includegraphics[width=.40\linewidth]{figures/host_test1_readwrite_no_temporal_hint.pdf}\label{fig:read_write}}

@@

\begin{figure*}
  \centering
  \subfigure[Performance of $y=Ax$ on the MIC. Squares: base 1/1 case,
    solid circles: 4/4 case implemented in C++, solid triangles: 4/4
    case implemented with MIC vector instructions. $-O3$ compilation
    options. Each symbol has two cases: with (red) and without (blue) 
    Reverse Cuthill-McKee bandwidth reduction. The grid is $96^3$.]
    {\includegraphics[width=.40\linewidth]{figures/mic_performance_nb_threads.pdf}\label{fig:perf_mic}}
%
\hspace{.4in}
%
  \subfigure[Performance on the host on a sparse matrix with $64^3$ rows. Manual
    vectorization brings significant improvement there as
    well.]{\includegraphics[width=.40\linewidth]{figures/64x64x64.pdf}\label{fig:perf_host}}
  \caption{Vectorization matters in practice. Differences in BW
    translate in different performance.\todo{this two plots should ``correlate'' }}
\end{figure*}

\begin{figure*}[t]
  \centering
  \subfigure[Supercompact and Random]{\includegraphics[width=.48\linewidth]{figures/random_supercompact.pdf}}
  \subfigure[Compact and RBF-FD]{\includegraphics[width=.48\linewidth]{figures/rbf_compact.pdf}}

  %\subfigure[Supercompact]{\includegraphics[width=.48\linewidth]{figures/supercompact_max_perf.pdf}}
  %\subfigure[Compact]{\includegraphics[width=.48\linewidth]{figures/compact_max_perf.pdf}} \\
  %\subfigure[Random]{\includegraphics[width=.48\linewidth]{figures/random_max_perf.pdf}}
  %\subfigure[RBF 3D]{\includegraphics[width=.48\linewidth]{figures/rbf_max_perf_run1.pdf}}
  \caption{Performance of "best code" accross matrix types. The two horizontal lines depict worst and 
  best case performance (Figure fig:gflops-peak-perf). RCM (blue, $\blacktriangle$) and no RCM (red,$\blacktriangledown$).}
\end{figure*}
